\sysfull

### 2018-08-06 paper/asplos19/paging.tex
typo
update algo 2
Paging revised (not complete)

% \textcolor{red}{[additional motivation for coala memory virtualization] we should emphasize that Coala programming model removes the burden of identifying the write-after-read dependencies from the programmer  as compared to Chain. And it does not require a specific compiler that cannot consider a runtime dependent behavior like Interrupts service handling to guarantee memory protection, as in Alpaca.}
%
\sys's {\em virtual memory manager} (VMM) abstracts the physical address space of the non-volatile memory (FRAM) and divides it into \texttt{\underline{private}} and \texttt{\underline{shadow}} (underlined locations are non-volatile) buffers. Each buffer is divided into pages. The VMM prohibits applications from directly access these buffers. Instead, it redirects any request to the \texttt{working} buffer located in the volatile memory (SRAM), which has a lower latency and energy cost to access than FRAM. It populates \texttt{working} buffer with the privatized pages from non-volatile memory. When a coalesced task completes, the VMM ensures that updated pages are committed to main memory and visible to subsequently executed tasks. If the power interrupts a task, the VMM ensures the consistency of the data by relying on its \texttt{\underline{private}} buffer. The VMM ensures consistency even if a task or a sequence of tasks include WAR dependences that could compromise consistency~\cite{ratchet,dino}.
%
On a commit, the VMM atomically copies all ``dirty'' pages back into their location in the non-volatile memory.
%
SRAM has a limited size and there is an extended working memory containing {\em
shadow pages} in the larger FRAM.  A shadow page stores the data from an
updated pages that need to be committed back to non-volatile memory, but that
exceeded the size of the volatile working memory.
%
the VMM moves pages of data efficiently using hardware accelerated Direct Memory
Access (DMA) support, which copies a block of memory from one place to another
as a single operation.\\
%
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{figures/mem-man.pdf}
    \caption{Interaction between task, memory manager and memory buffers upon \texttt{RP} and \texttt{WP},
    and order of occurrence. Steps 2, 3, 4 and 5 are conditional. Tasks only have direct access to the
    \texttt{working} buffer.}
    \label{figure:mem-man}
\end{figure}
%
\textbf{System Overview.}
Figure~\ref{figure:mem-man} gives an overview of \sys's virtual memory management system,
whose details are provided throughout this section.
A page has a home location in non-volatile memory which we refer to as the page's
\texttt{\underline{private}} (underlined locations are non-volatile).  When a
task accesses a page, the VMM swaps the page from its home location
into the fully\hyp{}associative \texttt{working} buffer that is stored in
volatile memory.  The \texttt{working} buffer is allocated to take up the
volatile memory that remains after space for stack is allocated.
%
The VMM utilizes a two-phase commit to write dirty pages from the \texttt{working}
buffer back to their home location in \texttt{\underline{private}}. It optimizes this process by using an indirection table to abstract the location of the \texttt{\underline{private}} buffer and
to eliminate the need for the second page copy.
%
The VMM is forced to swap out a dirty page from the \texttt{working} buffer
when a task requests a page that is not in the \texttt{working} buffer, and there is no
free page in the working buffer (i.e., a page fault). On a page fault, the VMM
moves the page to the direct\hyp{}mapped non-volatile \texttt{\underline{shadow}} buffer. \\
%

@@

%
A task must access protected non-volatile variables through \sys's
restricted memory interface. The interface includes \texttt{RP}$(v)$, to read the value of variable
$v$, and \texttt{WP}$(v)$ to assign a value to variable $v$. The
implementation of {\tt RP} is shown in Algorithm~\ref{algo:rwar}, and {\tt
WP}'s implementation is similar except that {\tt WP} marks the accessed page as
dirty.
%
{\tt RP} and {\tt WP} operations translate a variable's physical address in
non-volatile memory into a \emph{virtual address} in \texttt{working} buffer in
volatile memory. In \sys, a virtual address is composed of a \emph{page tag}
that identifies the page and a \emph{page offset} that identifies a byte.
%
\begin{figure}
\fbox{\parbox{\linewidth}{
	\captionof{algorithm}{\texttt{RP}(variable $v$)}
	\label{algo:rwar}
	\scriptsize
	\begin{algorithmic}[1]
		\State $t \leftarrow \Call{GetTag}{v}$
        \State $i \gets \{ j\ |\ \Call{GetTag}{\texttt{working}[j]} = t \}$ \Comment{Search page}
		\If { $i = \emptyset$ } \Comment{Var in a resident page?}
		\State	$i \gets \Call{PageFault}{t}$ \Comment{Swap page in}
		\EndIf
		\State $o \leftarrow \Call{GetOffset}{v}$
		\State \Return \texttt{working}[$i$][$o$]  \Comment{Return from page}
	\end{algorithmic}}}
\end{figure}
%
After address translation (Figure~\ref{figure:mem-man}, Step 6), a task
accesses the protected variable's location in the volatile \texttt{working} buffer
(Step 7). The VMM keeps track of the page tags for
the pages currently resident in the \texttt{working} buffer. When a task accesses a
variable, it compares the variable's page tag to tags of the pages in {\tt
working} (Algorithm~\ref{algo:rwar}, Line 2).
%
If the accessed variable's page tag is not found in the page buffer, the
operation incurs a {\em page fault} (Line 4).
%
The byte is accessed in the page buffer at the index of the resident page and
the variable's page offset (Lines 5--6).
%

@@

%
When accessing a protected variable with \texttt{RP} or \texttt{WP}, the memory
manager first searched the variable's page in the \texttt{working} buffer
(Figure~\ref{figure:mem-man}, Step 1).
If the page is not found there, a page fault is incurred and a new page needs to
be swapped in.
If the \texttt{working} buffer is full, a page fault on a memory access
requires VMM to swap out one of the pages in the \texttt{working} buffer (a
\emph{victim page}) preserving updates made to that page.
%
While we opted for the simplicity of FIFO, any other replacement policy, such
as Least Recently Used (LRU) could work in its place. \\
%
If a task modified any byte in the victim page (i.e., using \texttt{WP}), then
the page is dirty and its changes has to be persisted to non-volatile memory
(Figure~\ref{figure:mem-man}, Step 2).
If the accessed page was previously modified and swapped out since the last
power failure, the most recent version of the page is in {\tt
\underline{shadow}} (Step 3), otherwise it has to be retrieved from
{\tt \underline{private}} (Step 4).
Finally, the page is copied to the volatile buffer (Step 5). \\
%

@@

%
When the last task in a sequence of coalesced tasks completes, \sys must commit
\emph{all} dirty pages in the \texttt{working} buffer and in the
\texttt{\underline{shadow}} buffer, copying them back to their original
locations in the main memory \texttt{\underline{private}}.\\
To make the commit atomic, \sys commits the pages in two phases, implemented by
Algorithm~\ref{algo:commit}.
%
The first phase copies dirty pages from the \texttt{working} buffer to the
non-volatile \texttt{\underline{shadow}} buffer (Line 4). The second phase
commits pages from \texttt{\underline{shadow}} to \texttt{\underline{private}}
(Line 12).  If power fails during the first phase, the whole commit is aborted,
and the execution restarts from the most recently committed point (i.e., from the beginning of a coalesced task). If power fails in the second
phase, the commit process safely resumes on reboot.  The second phase of commit
depends on some run-time metadata. The \texttt{\underline{committing}} bit
indicates that a commit is in progress and is set before the first page is
committed (Line 9) and cleared after the last page is committed (Line 16).  The
\texttt{\underline{shadowCount}} records the number of dirty shadow pages to
commit and the VMM clears the counter when commit completes (Line 14).  The
\texttt{\underline{commitIndex}} indexes the next page to be committed (Line
11) and the VMM clears the index at the end of the phase (Line 15).
%
\sys's commit is efficient because it maintains an index dirty pages instead of
iterating through all potentially dirty pages to check a dirty bit during
commit.
%
\sys's implementation is also efficient because the VMM does not
copy from \texttt{\underline{shadow}} to \texttt{\underline{private}}, instead
swapping pointers in an indirection table that maintains the pages as a double
buffer.  Section~\ref{sec:implementation} describes double buffering in more
detail. \\
%
\textbf{Memory Consistency.} \sys's paging mechanism ensures that a task only
ever executes using consistent protected data. During task execution,
modifications to protected data do not affect the \texttt{\underline{private}} buffer,
because a task reads and writes the volatile \texttt{working} buffer only, and
modified pages are kept in \texttt{\underline{shadow}} until commit.  A power
failure erases the contents of the \texttt{working} buffer, preventing a re-executing task
from observing updates from a previous execution attempt.
Clearing \texttt{\underline{shadowCount}} as part of the second phase commit
(Line 14), ensures that all accesses to protected variables in subsequent tasks
correctly access their original memory locations in \texttt{\underline{private}}.\\
%
\begin{algorithm}[t]
	\caption{Two-phase commit}
	\label{algo:commit}
	\scriptsize
	\begin{algorithmic}[1]
        \Procedure{CommitPhase1}{} \Comment{On last coalesced task}
            \For {$i \in 0..|\texttt{working}|-1$}
                \State $t \gets \Call{Tag}{\texttt{working}[i]}$
                \State $\texttt{\underline{shadow}}[\Call{Tag}{\texttt{working}[i]}] \xleftarrow{\text{DMA}} \texttt{working}[i]$
                \State $\texttt{\underline{shadowList}}[\texttt{\underline{shadowCount}}] \gets t$
                \State $\texttt{\underline{shadowCount}} \gets \texttt{\underline{shadowCount}} + 1$
            \EndFor
            \State \Call{CommitPhase2}{}
        \EndProcedure
        \Procedure{CommitPhase2}{} \Comment{\texttt{\underline{commitIndex}} 0 on first boot}
            \State $\texttt{\underline{committing}} \gets \textbf{true}$
            \While{\texttt{\underline{commitIndex}} < \texttt{\underline{shadowCount}}}
                \State $t \gets \texttt{\underline{shadowList}}[\texttt{\underline{commitIndex}}]$
                \State \Call{commitToStore}{\texttt{\underline{shadow}}[t]}
                \State $\texttt{\underline{commitIndex}} \gets \texttt{\underline{commitIndex}} + 1$
            \EndWhile
            \State $\texttt{\underline{shadowCount}} \gets 0$
            \State $\texttt{\underline{commitIndex}} \gets 0$
            \State $\texttt{\underline{committing}} \gets \textbf{false}$
        \EndProcedure
        \Procedure{OnBoot}{} \Comment{Invoked on every boot}
            \If { \texttt{\underline{committing}} } \Call{CommitPhase2}{}
            \EndIf
            \State $\texttt{\underline{shadowCount}} \gets 0$, $i_\text{victim} \gets 0$
        \EndProcedure
	\end{algorithmic}
\end{algorithm}
%

@@

%
\sys asks the programmer to use its {\tt RP} and {\tt WP} paging API methods on
every access to a protected variable (cf. Section~\ref{sec:coala_api}).  These
API invocations present a risk of high overhead because there is a dynamic
check on every read and write. 
%
Despite the risk of per-access overhead, \sys's dynamic memory protection
scheme brings several benefits over a static approach (i.e.~\cite{alpaca}).  
First, the limitations of static analysis preclude some uses of pointers due to potential pointer
aliasing. For example, in the presence of arbitrary pointer operation, a
function call using a function pointer, or an interrupt within a task, the
system cannot statically analyze the memory behavior.   Second, a static approach
cannot handle task coalescing, because a protected variable's lifetime (i.e.,
from first use to commit) is unknown at compile time. \sys's dynamic,
per-access instrumentation supports arbitrary use of pointers and enables task
coalescing.

>>>>

%\note{additional motivation for coala memory virtualization] we should emphasize that Coala programming model removes the burden of identifying the write-after-read dependencies from the programmer  as compared to Chain. And it does not require a specific compiler that cannot consider a runtime dependent behavior like Interrupts service handling to guarantee memory protection, as in Alpaca.}

\sys's {\em virtual memory manager} is a core enabler of \sys task adaptation mechanism. Its overview is given in Figure~\ref{figure:mem-man}. VMM abstracts the physical address space of the non-volatile memory (FRAM) and divides it into \texttt{\underline{private}} and \texttt{\underline{shadow}} (underlined locations are non-volatile) buffers. Each buffer is divided into pages. The VMM prohibits applications from directly access these buffers. Instead, it redirects any request to the \texttt{working} buffer located in the volatile memory (SRAM), which has a lower latency and energy cost to access than FRAM. It populates \texttt{working} buffer with the privatized pages from non-volatile memory. When a coalesced task completes, the VMM ensures that updated pages are committed to main memory and visible to subsequently executed tasks. If the power interrupts a task, the VMM ensures the consistency of the data by relying on its \texttt{\underline{private}} buffer. The VMM ensures consistency even if a task or a sequence of tasks include WAR dependences that could compromise consistency~\cite{ratchet,dino}. On a commit, the VMM atomically copies, through two-phase commit, all ``dirty'' pages back into their location in the non-volatile memory. SRAM has a limited size and there is an extended working memory containing {\em shadow pages} in the larger FRAM. A shadow page stores the data from an updated pages that need to be committed back to non-volatile memory, but that exceeded the size of the volatile working memory. The VMM moves pages of data efficiently using hardware accelerated DMA support.

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{figures/mem-man.pdf}
    \caption{Interaction between task, memory manager and memory buffers upon \texttt{RP} and \texttt{WP}, and its order of occurrence; steps 2, 3, 4 and 5 are conditional. Tasks only have direct access to the \texttt{working} buffer.}
    \label{figure:mem-man}
\end{figure}

%\textbf{System Overview.} Figure~\ref{figure:mem-man} gives an overview of \sys's virtual memory management system. A page has a home location in non-volatile memory which we refer to as the page's \texttt{\underline{private}}. When a task accesses a page, the VMM swaps the page from its home location into the fully\hyp{}associative \texttt{working} buffer that is stored in volatile memory.  The \texttt{working} buffer is allocated to take up the volatile memory that remains after space for stack is allocated.

%The VMM utilizes a two-phase commit to write dirty pages from the \texttt{working} buffer back to their home location in \texttt{\underline{private}}. It optimizes this process by using an indirection table to abstract the location of the \texttt{\underline{private}} buffer and to eliminate the need for the second page copy. The VMM is forced to swap out a dirty page from the \texttt{working} buffer when a task requests a page that is not in the \texttt{working} buffer, and there is no free page in the working buffer (i.e., a page fault). On a page fault, the VMM moves the page to the direct\hyp{}mapped non-volatile \texttt{\underline{shadow}} buffer.

@@

A task must access protected non-volatile variables through \sys's restricted memory interface. The interface includes \texttt{RP}$(v)$, to read the value of variable $v$, and \texttt{WP}$(v)$ to assign a value to variable $v$. The implementation of {\tt RP} is shown in Algorithm~\ref{algo:rwar}, and {\tt WP}'s implementation is similar except that {\tt WP} marks the accessed page as dirty. {\tt RP} and {\tt WP} operations translate a variable's physical address in non-volatile memory into a \emph{virtual address} in \texttt{working} buffer in volatile memory. In \sys, a virtual address is composed of a \emph{page tag} that identifies the page and a \emph{page offset} that identifies a byte.

\begin{algorithm}[t]
    \captionof{algorithm}{\texttt{RP}(variable $v$)}
    \label{algo:rwar}
    \scriptsize
    \begin{algorithmic}[1]
		\State $t \leftarrow \Call{GetTag}{v}$
        \State $i \gets \{ j\ |\ \Call{GetTag}{\texttt{working}[j]} = t \}$ \Comment{Search page}
		\If { $i = \emptyset$ } \Comment{Var in a resident page?}
		\State	$i \gets \Call{PageFault}{t}$ \Comment{Swap page in}
		\EndIf
		\State $o \leftarrow \Call{GetOffset}{v}$
		\State \Return \texttt{working}[$i$][$o$]  \Comment{Return from page}
	\end{algorithmic}
\end{algorithm}

After address translation (Figure~\ref{figure:mem-man}, Step 6), a task accesses the protected variable's location in the volatile \texttt{working} buffer (Step 7). The VMM keeps track of the page tags for the pages currently resident in the \texttt{working} buffer. When a task accesses a variable, it compares the variable's page tag to tags of the pages in {\tt working} (Algorithm~\ref{algo:rwar}, Line 2). If the accessed variable's page tag is not found in the page buffer, the operation incurs a {\em page fault} (Line 4). The byte is accessed in the page buffer at the index of the resident page and the variable's page offset (Lines 5--6).

@@

When accessing a protected variable with \texttt{RP} or \texttt{WP}, the memory manager first searched the variable's page in the \texttt{working} buffer (Figure~\ref{figure:mem-man}, Step 1). If the page is not found there, a page fault is incurred and a new page needs to
be swapped in. If the \texttt{working} buffer is full, a page fault on a memory access requires VMM to swap out one of the pages in the \texttt{working} buffer (a \emph{victim page}) preserving updates made to that page. While we opted for the simplicity of FIFO, any other replacement policy, such as Least Recently Used (LRU) could work in its place. If a task modified any byte in the victim page (i.e., using \texttt{WP}), then the page is dirty and its changes has to be persisted to non-volatile memory (Figure~\ref{figure:mem-man}, Step 2). If the accessed page was previously modified and swapped out since the last power failure, the most recent version of the page is in {\tt \underline{shadow}} (Step 3), otherwise it has to be retrieved from {\tt \underline{private}} (Step 4). Finally, the page is copied to the volatile buffer (Step 5).

@@

When the last task in a sequence of coalesced tasks completes, \sys must commit \emph{all} dirty pages in the \texttt{working} buffer and in the \texttt{\underline{shadow}} buffer, copying them back to their original locations in the main memory \texttt{\underline{private}}.

To make the commit atomic, \sys commits the pages in two phases, implemented by Algorithm~\ref{algo:commit}. The first phase copies dirty pages from the \texttt{working} buffer to the non-volatile \texttt{\underline{shadow}} buffer (Line 4). The second phase commits pages from \texttt{\underline{shadow}} to \texttt{\underline{private}} (Line 12).  If power fails during the first phase, the whole commit is aborted, and the execution restarts from the most recently committed point (i.e., from the beginning of a coalesced task). If power fails in the second
phase, the commit process safely resumes on reboot.  The second phase of commit depends on some run-time metadata. The \texttt{\underline{committing}} bit indicates that a commit is in progress and is set before the first page is committed (Line 9) and cleared after the last page is committed (Line 16).  The \texttt{\underline{shadowCount}} records the number of dirty shadow pages to commit and the VMM clears the counter when commit completes (Line 14). The \texttt{\underline{commitIndex}} indexes the next page to be committed (Line 11) and the VMM clears the index at the end of the phase (Line 15).

\sys's commit is efficient because it maintains an index dirty pages instead of iterating through all potentially dirty pages to check a dirty bit during commit. \sys's implementation is also efficient because the VMM does not copy from \texttt{\underline{shadow}} to \texttt{\underline{private}}, instead swapping pointers in an indirection table that maintains the pages as a double buffer.  Section~\ref{sec:implementation} describes double buffering in more detail.

\textbf{Memory Consistency.} \sys's paging mechanism ensures that a task only ever executes using consistent protected data. During task execution, modifications to protected data do not affect the \texttt{\underline{private}} buffer, because a task reads and writes the volatile \texttt{working} buffer only, and modified pages are kept in \texttt{\underline{shadow}} until commit. A power failure erases the contents of the \texttt{working} buffer, preventing a re-executing task from observing updates from a previous execution attempt. Clearing \texttt{\underline{shadowCount}} as part of the second phase commit (Line 14), ensures that all accesses to protected variables in subsequent tasks correctly access their original memory locations in \texttt{\underline{private}}.

\begin{algorithm}[t]
	\caption{Two-phase commit}
	\label{algo:commit}
	\scriptsize
	\begin{algorithmic}[1]
        \Procedure{CommitPhase1}{} \Comment{On last coalesced task}
            \For {$i \in 0..|\texttt{working}|-1$}
                \State $t \gets \Call{Tag}{\texttt{working}[i]}$
                \State $\texttt{\underline{shadow}}[\Call{Tag}{\texttt{working}[i]}] \xleftarrow{\text{DMA}} \texttt{working}[i]$
                \State $\texttt{\underline{shadowList}}[\texttt{\underline{shadowCount}}] \gets t$
                \State $\texttt{\underline{shadowCount}} \gets \texttt{\underline{shadowCount}} + 1$
            \EndFor
            \State \Call{CommitPhase2}{}
        \EndProcedure
        \Procedure{CommitPhase2}{} \Comment{\texttt{\underline{commitIndex}} 0 on first boot}
            \State $\texttt{\underline{committing}} \gets \textbf{true}$
            \While{\texttt{\underline{commitIndex}} < \texttt{\underline{shadowCount}}}
                \State $t \gets \texttt{\underline{shadowList}}[\texttt{\underline{commitIndex}}]$
                \State \Call{commitToPrivate}{\texttt{\underline{shadow}}[t]}
                \State $\texttt{\underline{commitIndex}} \gets \texttt{\underline{commitIndex}} + 1$
            \EndWhile
            \State $\texttt{\underline{shadowCount}} \gets 0$
            \State $\texttt{\underline{commitIndex}} \gets 0$
            \State $\texttt{\underline{committing}} \gets \textbf{false}$
        \EndProcedure
        \Procedure{OnBoot}{} \Comment{Invoked on every boot}
            \If { \texttt{\underline{committing}} } \Call{CommitPhase2}{}
            \EndIf
            \State $\texttt{\underline{shadowCount}} \gets 0$, $i_\text{victim} \gets 0$
        \EndProcedure
	\end{algorithmic}
\end{algorithm}

@@

\sys asks the programmer to use its {\tt RP} and {\tt WP} paging API methods on every access to a protected variable (cf. Section~\ref{sec:coala_api}). These API invocations present a risk of high overhead because there is a dynamic check on every read and write. Despite the risk of per-access overhead, \sys's dynamic memory protection scheme brings several benefits over a static approach (i.e.~\cite{alpaca}). First, the limitations of static analysis preclude some uses of pointers due to potential pointer aliasing. For example, in the presence of arbitrary pointer operation, a function call using a function pointer, or an interrupt within a task, the system cannot statically analyze the memory behavior. Second, a static approach cannot handle task coalescing, because a protected variable's lifetime, i.e.,from first use to commit, is unknown at compile time. \sys's dynamic, per-access instrumentation supports arbitrary use of pointers and enables task coalescing.
